Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-15 ind_files/02_math_background.tex
Correct we/us to I/me for consistency

\subsection{Birth-death processes}
Mathematically, a birth-death process is a so-called `continuous-time Markov chain' in which only transitions between local states are allowed. In other words, a birth-death process is a stochastic process unfolding in continuous time such that
\begin{itemize}
	\item The process is `Markov', meaning that the future is statistically independent of the past given the present. In more mathematical terms, if the value of the stochastic process at time $t$ is given by $X_t$, $\mathbb{P}(\cdot | E)$ denotes probability conditioned on $E$, and $u < s \leq t$, then
	\begin{equation*}
	\mathbb{P}(X_t | X_s, X_u) = \mathbb{P}(X_t | X_s)
	\end{equation*}
	This equation is simply saying that if we have the information about the state of a process at time $s$, then we do not gain any more predictive power about the process at a future time $t$ if we have additional knowledge about the process from some past time $u < s$. A series of tosses of a fair coin is a simple example of a Markov process, since knowing whether a coin landed on heads during a previous toss does not change your predictive power about whether the coin will land on heads the next time you toss it.
	\item Direct transitions must be `local'. Mathematicians usually reserve the phrase `birth-death process' to processes that take values in the non-negative integers $\{0,1,2,3,4,\ldots\}$. In this case, only direct transitions from $n$ to $n \pm 1$ are allowed to occur. Biologically, this is saying that we observe the population on a fine enough timescale that the probability of two or more births/deaths occurring at the exact same time is very low and we can disallow it entirely in our models. The conditions for higher dimensional birth-death processes look similar.
\end{itemize}
Since these processes unfold in continuous time, they are characterized not by transition probabilities but by transition \emph{rates}, which can be thought of as the probability of transition `per unit time'. The quantity of interest is usually the probability of being in a particular state at a given point in time. The entire birth-death process can be described in terms of such a quantity, through a so-called `Master equation'. The master equation is a partial differential equation (PDE) for the probability of being in a given state at a given time, However, in all but the simplest cases, we can't actually solve this PDE, because it is simply too hard. The primary source of difficulty is non-linearity in the transition rates and the fact that transitions occur in discrete, discontinuous `jumps'. It is much easier to describe and analyze systems by using tools from stochastic calculus and partial differential equations, as we describe below.

\subsection{SDEs and the Fokker-Planck equation}\label{intro_SDE}
Stochastic systems which change continuously (in the state space) can be described in terms of a `stochastic differential equation' (SDE), which here is interchangeable with the phrase `It\^o process'. An SDE for a stochastic process $\{X_t\}_{t \geq 0}$ is an equation of the form
\begin{equation}
\label{ito_SDE_integral}
X_t = \int\limits_{0}^{t} F(s,X_s)ds + \int\limits_{0}^{t} G(s,X_s)dW_s
\end{equation}
where $F(t,x)$ and $G(t,x)$ are `nice' functions\footnote{For the mathematically oriented reader, there are two requirements: Firstly, we require the functions to have `linear growth', meaning that we can find a constant $C > 0$ such that $\|F(t,x)\| + \|G(t,x)\| \leq C(1+\|x\|)$ for every $x \in \mathbb{R}^{d}$ and $t > 0$. We also require `Lipschitz continuity', which means that we can find a constant $L > 0$ such that $\|F(t,x) - F(t,y)\| + \|G(t,x)-G(t,y)\| \leq L\|x-y\|$ for every pair $x,y \in \mathbb{R}^d$ and $t > 0$. Here, $\|\cdot\|$ denotes the natural norm on the space under consideration and for our cases will usually be the Euclidean norm. For biological systems, both of these conditions will usually be satisfied, and so we assume going further that all our SDEs are always well-defined and admit solutions.} In the physics literature, $F$ and $G$ are often called the `drift' and `diffusion' of the process respectively. However, we will not use this terminology here due to  potential confusion with genetic drift (which actually corresponds to the `diffusion' in the physics terminology). $W_t$ is the so-called `Wiener process' or `standard Brownian motion'. Named after the botanist Robert Brown (who was looking at the random erratic motion of pollen grains in water under a microscope), $\{W_t\}_{t \geq 0}$ is a stochastic process that is supposed to model `random noise' or `undirected diffusion' of a particle. If one imagines $W_t$ as recording the position of a small pollen grain at time $t$, then $W_t$ can be formally thought of as a process that has the following properties:
\begin{itemize}
	\item It starts at the origin, \emph{i.e} $W_0 = 0$. This is a harmless assumption made for convenience and amounts to a choice of coordinate system.
	\item It moves continuously, without sudden jumps across regions of space, \emph{i.e} the map $t \to W_t$ is continuous. This simply says that our pollen grain moves short distances in short intervals of time.
	\item The future movement is independent of past history. That is, given times $u < s < t$, the displacement $W_t - W_s$ is independent of the past position $W_u$.
	\item The movement is directionless and random, and displacement is normally distributed. More precisely, given two times $s < t$, the displacement $W_t - W_s$ follows a normal distribution with a mean of $0$ (this is the `directionless' part) and a variance of $t-s$ (this is the `random' part).
\end{itemize}
It can then be shown that since the motion is equally likely to be in any direction, the expected position at any point of time is the same as the initial position, \emph{i.e} $\mathbb{E}[W_t | W_0] = W_0 = 0$.

@@

Consider a population of identical individuals subject to some ecological rules that affect individuals' birth and death rates. Since all individuals are identical, we can only really track the population size through time. The population as a whole at any time $t$ can thus be characterized by a single number - its population size (Figure \ref{fig_1D_pop_description}). Imagine further that if a population has $n$ identical individuals, then, from the ecological rules, we can determine a \emph{birth rate} $b(n)$, which gives us a measure of the probability that a new individual will be born and the population size becomes $n+1$ `per unit time'. One must be slightly precise about what exactly they mean when they say `per unit time' since there are no discrete `time steps' for individuals to be born. Here, by `birth rate', we mean the probability that there will be a birth (and no death) per an \emph{infinitesimal} amount of time. More formally, letting $N_t$ denote the random variable representing the population size at time $t$ and letting $\mathbb{P}(E)$ denote the probability (in the common-sense usage) of an event $E$, the birth rate\footnote{Note that unlike usual ecology convention, this is \emph{not} a per-capita birth rate. The ecology per capita birth and death rates can be found by dividing my birth/death rates by the current population size $n(t)$} $b(n)$ of individuals a population with population size $n$ is the quantity
\begin{equation}
\label{1D_birthrate_defn}
b(n) \coloneqq \lim_{\epsilon \to 0}\frac{1}{\epsilon}\mathbb{P}\left(N_{t+\epsilon}=n+1 | N_{t}=n\right)
\end{equation}
Exactly analogously, we can also define the \emph{death rate} $d(n)$ of individuals in a population of $n$ individuals as the quantity
\begin{equation}
\label{1D_deathrate_defn}
d(n) \coloneqq \lim_{\epsilon \to 0}\frac{1}{\epsilon}\mathbb{P}\left(N_{t+\epsilon}=n-1 | N_{t}=n\right)
\end{equation}
An alternative, perhaps more intuitive characterization, of these same quantities is the following: If we have a population of size $n$, and we know that \emph{either a birth or a death} has just occurred, then, the probability that the event that occurred is a birth is
\begin{equation*}
\mathbb{P\big[\textrm{ birth } \big{|} \textrm{ something happened }\big]} = \frac{b(n)}{b(n)+d(n)}
\end{equation*}
and the probability that the event was instead a death is given by
\begin{equation*}
\mathbb{P\big[\textrm{ death } \big{|} \textrm{ something happened }\big]} = \frac{d(n)}{b(n)+d(n)}
\end{equation*}
\\

@@

\subsection{Stochastic fluctuations and the weak noise approximation}\label{sec_1D_WNA}
% The mathematics in this section is taken from~\cite{gardiner_stochastic_2009} with minor modifications\footnote{Gardiner uses a parameter $\epsilon>0$ such that noise increases with an increase in $\epsilon$. I have used a parameter $K>0$ such that noise decreases with an increase in $K$, which is more biologically relevant. Gardiner's exact results can be recovered by substituting $\epsilon = \frac{1}{\sqrt{K}}$}.
If we assume the noise is \emph{weak}, then we can go still further with analytic techniques by measuring fluctuations from the deterministic expectations, albeit with some slightly cumbersome calculations to arrive at the final expressions. We will grit our teeth and get through the algebra below, with my promise that the final answer is neat and easy to handle. It is clear that as $K \to \infty$, equation \eqref{1D_SDE} describes a deterministic process, obtained as the solution to
\begin{equation}
\label{1D_det_limit}
\frac{dx}{dt} = A^{-}(x) = b_{K}(x) - d_{K}(x)
\end{equation}
This is a very intuitive equation, saying that the rate of change of the population is equal to the birth rate minus the death rate. Let the solution of this equation  be given by $\alpha(t)$, so that $\frac{d{\alpha}}{dt}(t) = A^{-}(\alpha(t))$.

We can now measure (scaled) fluctuations from the deterministic solution $\alpha$ through a new variable $y=\sqrt{K}\left(x-\alpha(t)\right)$. For notational clarity, we will also introduce a new time variable $s=t$ which is equal to the original time variable (this is just so the equations look clearer). Let the probability density function of this new variable be given by $\Tilde{P}(y,s)$. In summary, we have introduced the variables:
\begin{align*}
y &= \sqrt{K}\left(x-\alpha(t)\right)\\
s &= t\\
\Tilde{P}(y,s) &= \frac{1}{\sqrt{K}}P(x,t)
\end{align*}
Note that by ordinary rules of variable substitution, we have:
\begin{align}
\frac{\partial \Tilde{P}}{\partial t} &= \frac{\partial \Tilde{P}}{\partial y}\frac{\partial y}{\partial t} + \frac{\partial \Tilde{P}}{\partial s}\frac{\partial s}{\partial t}\nonumber\\
&=\frac{\partial \Tilde{P}}{\partial y}\left( -\sqrt{K}\frac{d\alpha}{dt}\right) + \frac{\partial \Tilde{P}}{\partial s}\nonumber\\
&= -\sqrt{K}A^{-}(\alpha(s))\frac{\partial \Tilde{P}}{\partial y} + \frac{\partial \Tilde{P}}{\partial s}\label{weak_noise_expansion_first_term}
\end{align}
and
\begin{equation}
\label{weak_noise_expansion_second_term}
\frac{\partial }{\partial y} = \frac{1}{\sqrt{K}}\frac{\partial }{\partial x}
\end{equation}
Reformulating \eqref{1D_FPE} in terms of $y,s$ and $\Tilde{P}$ and substituting \eqref{weak_noise_expansion_first_term} and \eqref{weak_noise_expansion_second_term} yields:
\begin{align}
-A^-(\alpha)\frac{\partial\Tilde{P}}{\partial x} + \frac{\partial \Tilde{P}}{\partial s} &= -\sqrt{K}\frac{\partial}{\partial y}\left(A^-(\alpha+\frac{y}{\sqrt{K}})\Tilde{P}\right)+\frac{1}{2}\frac{\partial^2}{\partial y^2}\left(A^+(\alpha+\frac{y}{\sqrt{K}})\Tilde{P}\right)\nonumber\\
\Rightarrow \frac{\partial \Tilde{P}}{\partial s} &= -\frac{\partial}{\partial y}\left[\sqrt{K}\left(A^-(\alpha+\frac{y}{\sqrt{K}})-A^-(\alpha)\right)\Tilde{P}\right]+\frac{1}{2}\frac{\partial^2}{\partial y^2}\left(A^+(\alpha+\frac{y}{\sqrt{K}})\Tilde{P}\right)\label{weak_noise_exact_equation}
\end{align}
We are now ready to make a weak noise `expansion'. We do so by assuming that $\Tilde{P}$, $A^-(\alpha + \frac{y}{\sqrt{K}})$, and $A^+(\alpha+\frac{y}{\sqrt{K}})$ can be approximated by series expansions in $\frac{1}{\sqrt{K}}$ of the form:
\begin{align*}
\Tilde{P} &= \sum\limits_{n=0}^{\infty}\Tilde{P}_n\left(\frac{1}{\sqrt{K}}\right)^n\\
A^-\left(\alpha(s) + \frac{y}{\sqrt{K}}\right) &= \sum\limits_{n=0}^{\infty}A^-_n(s)\left(\frac{y}{\sqrt{K}}\right)^n\\
A^+\left(\alpha(s) + \frac{y}{\sqrt{K}}\right) &= \sum\limits_{n=0}^{\infty}A^+_n(s)\left(\frac{y}{\sqrt{K}}\right)^n
\end{align*}
with $A^-_0(s) = A^-(\alpha(s)), A^+_0(s) = A^+(\alpha(s))$. These could be Taylor expansions, for example, but the exact form of the coefficients is irrelevant as long as it is known to us, so any expansion will work. We can now substitute these series expansions into \eqref{weak_noise_exact_equation} to obtain:
\begin{equation}
\begin{split}
\label{weak_noise_full_series_expansion}
\sum\limits_{n=0}^{\infty}\left(\frac{1}{\sqrt{K}}\right)^n\frac{\partial \Tilde{P}_n}{\partial s} = 
-\frac{\partial}{\partial y}\left[\sqrt{K}\left( \sum\limits_{n=1}^{\infty}A^-_n(s)\left(\frac{y}{\sqrt{K}}\right)^n\right)\left(\sum\limits_{m=0}^{\infty}\Tilde{P}_m\left(\frac{1}{\sqrt{K}}\right)^m\right)\right] \\ + 
\frac{1}{2}\frac{\partial^2}{\partial y^2}\left[\left(\sum\limits_{n=0}^{\infty}A^+_n(s)\left(\frac{y}{\sqrt{K}}\right)^n\right)\left(\sum\limits_{m=0}^{\infty}\Tilde{P}_m\left(\frac{1}{\sqrt{K}}\right)^m\right)\right]
\end{split}
\end{equation}
We can now compare the coefficients of $K^{-n/2}$ for each $n$ in order to arrive at approximations in the series expansion, the idea being that you neglect all terms which are of order greater than $\mathcal{O}(K^{-m/2})$ for some $m$ according to the desired precision.

>>>>

\subsection{Birth-death processes}
Mathematically, a birth-death process is a so-called `continuous-time Markov chain' in which only transitions between local states are allowed. In other words, a birth-death process is a stochastic process unfolding in continuous time such that
\begin{itemize}
	\item The process is `Markov', meaning that the future is statistically independent of the past given the present. In more mathematical terms, if the value of the stochastic process at time $t$ is given by $X_t$, $\mathbb{P}(\cdot | E)$ denotes probability conditioned on $E$, and $u < s \leq t$, then
	\begin{equation*}
	\mathbb{P}(X_t | X_s, X_u) = \mathbb{P}(X_t | X_s)
	\end{equation*}
	This equation is simply saying that if we have the information about the state of a process at time $s$, then we do not gain any more predictive power about the process at a future time $t$ if we have additional knowledge about the process from some past time $u < s$. A series of tosses of a fair coin is a simple example of a Markov process, since knowing whether a coin landed on heads during a previous toss does not change your predictive power about whether the coin will land on heads the next time you toss it.
	\item Direct transitions must be `local'. In one dimension, mathematicians usually reserve the phrase `birth-death process' to processes that take values in the non-negative integers $\{0,1,2,3,4,\ldots\}$. In this case, only direct transitions from $n$ to $n \pm 1$ are allowed to occur. Biologically, this is saying that we observe the population on a fine enough timescale that the probability of two or more births/deaths occurring at the exact same time is very low and we can disallow it entirely in our models. The conditions for higher dimensional birth-death processes look similar.
\end{itemize}
Since these processes unfold in continuous time, they are characterized not by transition probabilities but by transition \emph{rates}, which can be thought of as the probability of transition `per unit time'. The quantity of interest is usually the probability of being in a particular state at a given point in time. The entire birth-death process can be described in terms of such a quantity, through a so-called `Master equation'. The master equation is a partial differential equation (PDE) for the probability of being in a given state at a given time, However, in all but the simplest cases, we can't actually solve this PDE, because it is simply too hard. The primary source of difficulty is non-linearity in the transition rates and the fact that transitions occur in discrete, discontinuous `jumps'. It is much easier to describe and analyze systems that change `continuously', since one can then use tools from stochastic calculus and partial differential equations, as I describe below.

\subsection{SDEs and the Fokker-Planck equation}\label{intro_SDE}
Stochastic systems which change continuously (in the state space) can be described in terms of a `stochastic differential equation' (SDE), which here is interchangeable with the phrase `It\^o process'. An SDE for a stochastic process $\{X_t\}_{t \geq 0}$ is an equation of the form
\begin{equation}
\label{ito_SDE_integral}
X_t = \int\limits_{0}^{t} F(s,X_s)ds + \int\limits_{0}^{t} G(s,X_s)dW_s
\end{equation}
where $F(t,x)$ and $G(t,x)$ are `nice' functions\footnote{For the mathematically oriented reader, there are two requirements: Firstly, we require the functions to have `linear growth', meaning that we can find a constant $C > 0$ such that $\|F(t,x)\| + \|G(t,x)\| \leq C(1+\|x\|)$ for every $x \in \mathbb{R}^{d}$ and $t > 0$. We also require `Lipschitz continuity', which means that we can find a constant $L > 0$ such that $\|F(t,x) - F(t,y)\| + \|G(t,x)-G(t,y)\| \leq L\|x-y\|$ for every pair $x,y \in \mathbb{R}^d$ and $t > 0$. Here, $\|\cdot\|$ denotes the natural norm on the space under consideration and for our cases will usually be the Euclidean norm. For biological systems, both of these conditions will usually be satisfied, and so I assume going further that all our SDEs are always well-defined and admit solutions.} In the more mathematical (and physical) literature, $F$ and $G$ are often called the `drift' and `diffusion' of the process respectively. However, I will not use this terminology here to avoid potential confusion with genetic drift (which actually corresponds to the `diffusion' term in the math/physics terminology). $W_t$ is the so-called `Wiener process' or `standard Brownian motion'. Named after the botanist Robert Brown (who was looking at the random erratic motion of pollen grains in water under a microscope), $\{W_t\}_{t \geq 0}$ is a stochastic process that is supposed to model `random noise' or `undirected diffusion' of a particle. If one imagines $W_t$ as recording the position of a small pollen grain at time $t$, then $W_t$ can be formally thought of as a process that has the following properties:
\begin{itemize}
	\item It starts at the origin, \emph{i.e} $W_0 = 0$. This is a harmless assumption made for convenience and amounts to a choice of coordinate system.
	\item It moves continuously, without sudden jumps across regions of space, \emph{i.e} the map $t \to W_t$ is continuous. This simply says that our pollen grain moves short distances in short intervals of time.
	\item The future movement is independent of past history. That is, given times $u < s < t$, the displacement $W_t - W_s$ is independent of the past position $W_u$.
	\item The movement is directionless and random, and displacement is normally distributed. More precisely, given two times $s < t$, the displacement $W_t - W_s$ follows a normal distribution with a mean of $0$ (this is the `directionless' part) and a variance of $t-s$ (this is the `random' part).
\end{itemize}
It can then be shown that since the motion is equally likely to be in any direction, the expected position at any point of time is the same as the initial position, \emph{i.e} $\mathbb{E}[W_t | W_0] = W_0 = 0$.

@@

Consider a population of identical individuals subject to some ecological rules that affect individuals' birth and death rates. I will neglect any potential factor that could lead to two individuals becoming non-identical, like mutations, etc. Since all individuals are identical, we only really need to track the total population size through time to know everything there is to know about the population. The population as a whole at any time $t$ can thus be characterized by a single number - its population size (Figure \ref{fig_1D_pop_description}). Imagine further that if a population has $n$ identical individuals, then, from the ecological rules, we can determine a \emph{birth rate} $b(n)$, which gives us a measure of the probability that a new individual will be born and the population size becomes $n+1$ `per unit time'. One must be slightly precise about what exactly they mean when they say `per unit time' since there are no discrete `time steps' for individuals to be born. Here, by `birth rate', I mean the probability that there will be a birth (and no death) per an \emph{infinitesimal} amount of time. More formally, letting $N_t$ denote the random variable representing the population size at time $t$ and letting $\mathbb{P}(E)$ denote the probability (in the common-sense usage) of an event $E$, the birth rate\footnote{Note that unlike usual ecology convention, this is \emph{not} a per-capita birth rate. The ecology per capita birth and death rates can be found by dividing my birth/death rates by the current population size $n(t)$} $b(n)$ of individuals a population with population size $n$ is the quantity
\begin{equation}
\label{1D_birthrate_defn}
b(n) \coloneqq \lim_{\epsilon \to 0}\frac{1}{\epsilon}\mathbb{P}\left(N_{t+\epsilon}=n+1 | N_{t}=n\right)
\end{equation}
Exactly analogously, we can also define the \emph{death rate} $d(n)$ of individuals in a population of $n$ individuals as the quantity
\begin{equation}
\label{1D_deathrate_defn}
d(n) \coloneqq \lim_{\epsilon \to 0}\frac{1}{\epsilon}\mathbb{P}\left(N_{t+\epsilon}=n-1 | N_{t}=n\right)
\end{equation}
An alternative, perhaps more intuitive characterization, of these same quantities is the following: If we have a population of size $n$, and we know that \emph{either a birth or a death} has just occurred, then, the probability that the event that occurred is a birth is
\begin{equation*}
\mathbb{P\big[\textrm{ birth } \big{|} \textrm{ something happened }\big]} = \frac{b(n)}{b(n)+d(n)}
\end{equation*}
and the probability that the event was instead a death is given by
\begin{equation*}
\mathbb{P\big[\textrm{ death } \big{|} \textrm{ something happened }\big]} = \frac{d(n)}{b(n)+d(n)}
\end{equation*}
\\

@@

\subsection{Stochastic fluctuations and the weak noise approximation}\label{sec_1D_WNA}
% The mathematics in this section is taken from~\cite{gardiner_stochastic_2009} with minor modifications\footnote{Gardiner uses a parameter $\epsilon>0$ such that noise increases with an increase in $\epsilon$. I have used a parameter $K>0$ such that noise decreases with an increase in $K$, which is more biologically relevant. Gardiner's exact results can be recovered by substituting $\epsilon = \frac{1}{\sqrt{K}}$}.
If we assume the noise is weak (this will be made precise shortly), then we can go still further with analytic techniques by measuring fluctuations from the deterministic expectations, albeit with some slightly cumbersome calculations to arrive at the final expressions. We will grit our teeth and get through the algebra below, with my promise that the final answer is neat and easy to handle. It is clear that as $K \to \infty$, equation \eqref{1D_SDE} describes a deterministic process, obtained as the solution to
\begin{equation}
\label{1D_det_limit}
\frac{dx}{dt} = A^{-}(x) = b_{K}(x) - d_{K}(x)
\end{equation}
This is a very intuitive equation, saying that the rate of change of the population is equal to the birth rate minus the death rate. Let the solution of this equation  be given by $\alpha(t)$, so that $\frac{d{\alpha}}{dt}(t) = A^{-}(\alpha(t))$.

We can now measure (scaled) fluctuations from the deterministic solution $\alpha$ through a new variable $y=\sqrt{K}\left(x-\alpha(t)\right)$. For notational clarity, I will also introduce a new time variable $s=t$ which is equal to the original time variable (this is just so the equations look clearer). Let the probability density function of this new variable be given by $\Tilde{P}(y,s)$. In summary, I have introduced the variables:
\begin{align*}
y &= \sqrt{K}\left(x-\alpha(t)\right)\\
s &= t\\
\Tilde{P}(y,s) &= \frac{1}{\sqrt{K}}P(x,t)
\end{align*}
Note that by ordinary rules of variable substitution, we have:
\begin{align}
\frac{\partial \Tilde{P}}{\partial t} &= \frac{\partial \Tilde{P}}{\partial y}\frac{\partial y}{\partial t} + \frac{\partial \Tilde{P}}{\partial s}\frac{\partial s}{\partial t}\nonumber\\
&=\frac{\partial \Tilde{P}}{\partial y}\left( -\sqrt{K}\frac{d\alpha}{dt}\right) + \frac{\partial \Tilde{P}}{\partial s}\nonumber\\
&= -\sqrt{K}A^{-}(\alpha(s))\frac{\partial \Tilde{P}}{\partial y} + \frac{\partial \Tilde{P}}{\partial s}\label{weak_noise_expansion_first_term}
\end{align}
and
\label{weak_noise_expansion_second_term}
\frac{\partial }{\partial y} = \frac{1}{\sqrt{K}}\frac{\partial }{\partial x}
Reformulating \eqref{1D_FPE} in terms of $y,s$ and $\Tilde{P}$ and substituting \eqref{weak_noise_expansion_first_term} and \eqref{weak_noise_expansion_second_term} yields:
\begin{align}
-A^-(\alpha)\frac{\partial\Tilde{P}}{\partial x} + \frac{\partial \Tilde{P}}{\partial s} &= -\sqrt{K}\frac{\partial}{\partial y}\left(A^-(\alpha+\frac{y}{\sqrt{K}})\Tilde{P}\right)+\frac{1}{2}\frac{\partial^2}{\partial y^2}\left(A^+(\alpha+\frac{y}{\sqrt{K}})\Tilde{P}\right)\nonumber\\
\Rightarrow \frac{\partial \Tilde{P}}{\partial s} &= -\frac{\partial}{\partial y}\left[\sqrt{K}\left(A^-(\alpha+\frac{y}{\sqrt{K}})-A^-(\alpha)\right)\Tilde{P}\right]+\frac{1}{2}\frac{\partial^2}{\partial y^2}\left(A^+(\alpha+\frac{y}{\sqrt{K}})\Tilde{P}\right)\label{weak_noise_exact_equation}
\end{align}
We are now ready to make a weak noise `expansion' (This is a special case of a more general idea called a `perturbative expansion' or `perturbation theory' in physics and related fields). We do so by assuming that $\Tilde{P}$, $A^-(\alpha + \frac{y}{\sqrt{K}})$, and $A^+(\alpha+\frac{y}{\sqrt{K}})$ can be approximated by series expansions in $\frac{1}{\sqrt{K}}$ of the form:
\begin{align*}
\Tilde{P} &= \sum\limits_{n=0}^{\infty}\Tilde{P}_n\left(\frac{1}{\sqrt{K}}\right)^n\\
A^-\left(\alpha(s) + \frac{y}{\sqrt{K}}\right) &= \sum\limits_{n=0}^{\infty}A^-_n(s)\left(\frac{y}{\sqrt{K}}\right)^n\\
A^+\left(\alpha(s) + \frac{y}{\sqrt{K}}\right) &= \sum\limits_{n=0}^{\infty}A^+_n(s)\left(\frac{y}{\sqrt{K}}\right)^n
\end{align*}
with $A^-_0(s) = A^-(\alpha(s)), A^+_0(s) = A^+(\alpha(s))$. These could be Taylor expansions, for example, but the exact form of the coefficients is irrelevant as long as it is known to us, so any expansion will work. We can now substitute these series expansions into \eqref{weak_noise_exact_equation} to obtain:
\begin{split}
\label{weak_noise_full_series_expansion}
\sum\limits_{n=0}^{\infty}\left(\frac{1}{\sqrt{K}}\right)^n\frac{\partial \Tilde{P}_n}{\partial s} = 
-\frac{\partial}{\partial y}\left[\sqrt{K}\left( \sum\limits_{n=1}^{\infty}A^-_n(s)\left(\frac{y}{\sqrt{K}}\right)^n\right)\left(\sum\limits_{m=0}^{\infty}\Tilde{P}_m\left(\frac{1}{\sqrt{K}}\right)^m\right)\right] \\ + 
\frac{1}{2}\frac{\partial^2}{\partial y^2}\left[\left(\sum\limits_{n=0}^{\infty}A^+_n(s)\left(\frac{y}{\sqrt{K}}\right)^n\right)\left(\sum\limits_{m=0}^{\infty}\Tilde{P}_m\left(\frac{1}{\sqrt{K}}\right)^m\right)\right]
\end{split}
We can now compare the coefficients of $K^{-n/2}$ for each $n$ in order to arrive at approximations in the series expansion, the idea being that you neglect all terms which are of order greater than $\mathcal{O}(K^{-m/2})$ for some $m$ according to the desired precision.

We observe that for any fixed $r$, the coefficient of $K^{-r/2}$ on the LHS is $\frac{\partial \Tilde{P}_r}{\partial s}$. On the RHS, the coefficients of $K^{-r/2}$ in the second term have the form $\Tilde{P}_{m}A^+_{n}y^n$, subject to the constraint that $m+n=r$. Furthermore, all such terms (and only such terms) are coefficients of $K^{-r/2}$. Thus, after grouping, the coefficient of $K^{-r/2}$ from the second terms of the RHS of \eqref{weak_noise_full_series_expansion} is precisely
\begin{equation*}
\frac{1}{2}\frac{\partial^2}{\partial y^2}\sum\limits_{m=0}^{r}\Tilde{P}_{m}A^+_{r-m}y^{r-m}
\end{equation*}
Exactly analogous reasoning reveals that the contribution of the first term of the RHS is:
\begin{equation*}
-\frac{\partial}{\partial y}\sum\limits_{m=0}^{r}\Tilde{P}_{m}A^-_{r-m+1}y^{r-m+1}
\end{equation*}
Thus, we find that the $r$th term of the expansion satisfies:
\begin{equation}
\label{weak_noise_expansion_each_term}
\frac{\partial \Tilde{P}_r}{\partial s} = -\frac{\partial}{\partial y}\left(\sum\limits_{m=0}^{r}\Tilde{P}_{m}A^-_{r-m+1}y^{r-m+1}\right) + \frac{1}{2}\frac{\partial^2}{\partial y^2}\left(\sum\limits_{m=0}^{r}\Tilde{P}_{m}A^+_{r-m}y^{r-m}\right)
\end{equation}
We will now make the meaning of `weak' precise. Let us assume that the fluctuations are weak enough that we can obtain a reasonable approximation of the dynamics by retaining only the highest order (in $1/K$) term in equation \eqref{weak_noise_expansion_each_term} and neglecting all higher-order terms\footnote{For example, we may imagine this is reasonable if the deterministic trajectory is at a stable fixed point and subject to weak fluctuations}. We are then left with the expression:
\begin{equation}
\label{1D_WNA}
\frac{\partial \Tilde{P}_0}{\partial s} = -A^-_1(s)\frac{\partial}{\partial y}(y\Tilde{P}_0) + \frac{A^+_{0}(s)}{2}\frac{\partial^2 \Tilde{P}_{0}}{\partial y^2}
\end{equation}
which is simply the Fokker-Planck equation for the It\^{o} process
\begin{equation*}
dY_t = A^-_1(t)Y_tdt + \sqrt{A^+_0(t)}dW_t
\end{equation*}
This is the `weak noise approximation', (sometimes also called the `linear noise approximation' because the resulting Fokker-Planck equation is linear). This equation describes a so-called `Ornstein-Uhlenbeck process', and is easily solved by using $\exp(-\int A^-_1(s)ds)$ as an `integrating factor'. In particular, multiplying both sides by $\exp(-\int A^-_1(s)ds)$ yields
\begin{align*}
\exp\left(-{\int\limits_{0}^{t}A^-_1(s)ds}\right)dY_t - Y_tA^-_1(t)\exp\left(-{\int\limits_{0}^{t}A^-_1(s)ds}\right)dt &= \sqrt{A^+_0(t)}\exp\left(-{\int\limits_{0}^{t}A^-_1(s)ds}\right)dW_t\\
\Rightarrow d\left(\exp\left(-{\int\limits_{0}^{t}A^-_1(s)ds}\right)Y_t\right) &= \sqrt{A^+_0(t)}\exp\left(-{\int\limits_{0}^{t}A^-_1(s)ds}\right)dW_t
\end{align*}
Integrating both sides and noting that $A^+_0(s) = A^+(\alpha(s))$, we thus obtain the final expression
\begin{equation}
\label{weak_noise_OU_solution}
Y_t = Y_0\exp\left({\int\limits_{0}^{t}A^-_1(s)ds}\right)+\int\limits_{0}^{t}\exp\left(-\int\limits_{s}^{t}A^-_{1}(v)dv\right)\sqrt{A^+(\alpha(s))}dW_s
\end{equation}
as the zeroth-order weak noise approximation for stochastic fluctuations from the deterministic trajectory due to demographic noise Note that this is an exact equation, and one can get many insights from it. For example, if $Y_0 = 0$ (\emph{i.e} we start at the deterministic steady state, a natural assumption for measuring fluctuations from it), then we can show by taking expectations in \eqref{weak_noise_OU_solution} and using results presented in \ref{intro_SDE} that we must have $\mathbb{E}[Y_t | Y_0] = 0$. In other words, the fluctuations have zero expectation and are expected to occur symmetrically about $\alpha(t)$), with no bias. The variance (spread) of the fluctuations $Y_t$, as well as higher moments, can also be exactly calculated from \eqref{weak_noise_OU_solution} using some tools from stochastic calculus, but I will not demonstrate this here.
